NMRlipids III: Lipid-cholesterol interactions in atomistic resolution molecular dynamics simulations

### 2019-05-08 MANUSCRIPT/manuscript.tex
Update manuscript

@@

Addition of cholesterol substantially increases the acyl chain C-H order parameters in PC lipid bilayers (Fig. \ref{OrderParametersCHOL}),
which can be explained by increased trans conformations in acyl chains \cite{ferreira13,??}.
%Acyl chain order parameters for pure 1-palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
%and mixture with 50 mol\% of cholesterol from different simulations and experiments
%are shown in Fig. \ref{OrderParametersCHOL}.
The absolute values of order parameters of sn-1 chain from $^{13}$C NMR experiments \cite{ferreira13}
exhibit approximately linear increase with cholesterol up to equimolar mixture (Fig. \ref{slopes}), which is expected because
phase separation is not observed for this mixture \cite{ionova12,ferreira13}.
%thus the intermolecular interactions are expected to have maximum effect in equimolar mixture.
\todo{Is overestimation in CHARMM36 and Slipid significant or not?} \\
\todo{Lipid14 results from recent article should be added in this figure} 
\begin{figure*}[]
  \centering
  \includegraphics[width=17.2cm]{../FIGS/OrderParametersCHOL.eps}
  \caption{\label{OrderParametersCHOL}
    Order parameters from simulations and experiments for acyl chains of  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).
  }
\end{figure*}

Experimental acyl chain order parameters of pure POPC lipid bilayers
are well reproduced by most force fields (Fig. \ref{OrderParametersCHOL}),
except for the C$_2$ carbon in {\it sn}-1 chain,
which is typically the case in the state of the art lipid force fields (for review see \cite{ollila16}).
%This is also mainly observed in the data
%presented here, except that
However, the order parameters in the beginning of {\it sn}-1 chain are underestimated in the
MACROG simulation and CHARMM36 simulations ran also with other than Gromacs
(section \ref{CHARMMprograms} in the supplementary information) slightly overestimates the order parameters.
The absolute values of {\it sn}-1 acyl chain order parameters exhibit approximately linear increase upon addition
of cholesterol also in simulations below equimolar mixture (Figs. \ref{slopesberger}-\ref{slopesslipids310K}).
However, the slope of the increase (Fig. \ref{OrderParametersCHOLchanges}) and order parameters
in equimolar mixtures (Fig. \ref{OrderParametersCHOL}) are overestimated by all force fields.
%For Berger/H\"oltje and MACROG models the overestimation is significant,
%while less severe for CHARMM36 and Slipid models.
The overestimation of order parameters observed in CHARMM without cholesterol and in all force fields
with cholesterol is smaller than the contribution by undulations in large simulations (section \ref{undulations} in the supplementary information).
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/slopes.eps}
  \caption{\label{OrderParametersCHOLchanges}
    Slopes of order parameters as a function of cholesterol.
  }
\end{figure}

@@

\subsection{Lipid bilayer dimensions and density profiles as a function of cholesterol}
Cholesterol induced changes in lipid packing, thickness and density profiles are suggested
to be relevant, for example, for membrane protein interactions and cellular membrane
lateral organization \cite{??}. Experimental and simulation studies show that
the cholesterol induces the so called ''condensing effect'', i.e. acyl chain ordering
increases the membrane thickness and
reduces the area per molecule in lipid bilayers \cite{??}. This is demonstrated
from simulations in Fig. \ref{apls}, showing the area per PC headgroups and per
total number of molecules as a function of cholesterol concentration from different
models. All models show reduction of area per total amount of molecules, partly
due to the smaller area covered by the cholesterol than lipids, but also because
lipids ordered by cholesterol require less space. Due to the latter effect,
the area per PC headgroup does not essentially increase up to the addition of
$\sim$15 mol\% of cholesterol. \\
\todo{Membrane thickness to be plotted as well.}
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/apls.eps}
  \caption{\label{apls}
    Area per molecules calculated from different simulation models
    as a function of cholesterol concentration. The solid symbols are area per
    total amount of molecules (chol+PC) and the empty symbols area per PC headgroups.
    Top figure shows absolute values and bottom figure shows changes respect
    to pure lipid system.
  }
\end{figure}

Bilayer dimensions are indirectly experimentally accessible by scattering methods 
and have been widely used to validate lipid bilayer simulations \cite{ollila16,??}. 
While sophisticated models are applied to extract area per molecule and thicknes for single
component lipid bilayers, the results for multicomponent systems are more difficult to interpret \cite{pan12,Heftberger15,Marquardt15,??}.
In such case the validation can be done by comparing directly measurable form factors between
simulations and experiments. If simulation reproduces the form factor, it can be also
used to interpret the experiments.  The form factors from different simulation models with
different cholesterol concentrations are compared to experimental data in Fig.~\ref{FormFactors}. \\
\todo{Discussion to be finished when the data is solid.}

 \begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/FormFactors.eps}
  \caption{\label{FormFactors}
    Form factors from simulations and experiments.
  }
  \todo{Form factor calculation method should be double checked.}
  \todo{Experimental form factor amplitudes are not scaled to match with simulations, as done usually}
  \todo{Not all experimental and simulation data is here.}
\end{figure}

 \begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/FFfitness.eps}
  \caption{\label{FormFactorsFITNESS}
    Fitness of form factors between simulations and experiments.
  }
 \end{figure}

  \begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/densitiesEXTREMES.eps}
  \caption{\label{densities}
    Density profiles from simulations and experiments.
  }
\end{figure}

@@

@@

Cholesterol ordering effect is overestimated in Berger/Holtje and MacRog models.
Slight overestimation is observed also in CHARMM36 and Slipid models, but more
careful analysis is required to conclude if this is significant or not.

>>>>

\clearpage

@@

Addition of cholesterol substantially increases the acyl chain C-H order parameters in PC lipid bilayers (Fig. \ref{OrderParametersCHOL}),
which can be explained by increased trans conformations in acyl chains \cite{ferreira13,??}.
%Acyl chain order parameters for pure 1-palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
%and mixture with 50 mol\% of cholesterol from different simulations and experiments
%are shown in Fig. \ref{OrderParametersCHOL}.
The absolute values of order parameters of sn-1 chain from $^{13}$C NMR experiments \cite{ferreira13}
exhibit approximately linear increase with cholesterol up to equimolar mixture (Fig. \ref{slopes}), which is expected because
phase separation is not observed for this mixture \cite{ionova12,ferreira13}.
%thus the intermolecular interactions are expected to have maximum effect in equimolar mixture.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.2cm]{../FIGS/OrderParametersCHOL.eps}
  \caption{\label{OrderParametersCHOL}
    Order parameters from simulations and experiments for acyl chains of  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).
  }
  \todo{Lipid14 results?} 
\end{figure*}

Experimental acyl chain order parameters of pure POPC lipid bilayers
are well reproduced by most force fields (Fig. \ref{OrderParametersCHOL}),
except for the C$_2$ carbon in {\it sn}-1 chain,
which is typically the case in the state of the art lipid force fields (for review see \cite{ollila16}).
%This is also mainly observed in the data
%presented here, except that
However, the order parameters in the beginning of {\it sn}-1 chain are underestimated in the
MACROG simulation and CHARMM36 simulations slightly overestimates the order parameters, also when
used with other than Gromacs simulation packages (section \ref{CHARMMprograms} in the supplementary information).
The absolute values of {\it sn}-1 acyl chain order parameters exhibit approximately linear increase upon addition
of cholesterol also in simulations below equimolar mixture (Figs. \ref{slopesberger}-\ref{slopesslipids310K}).
However, the slope of the increase (Fig. \ref{OrderParametersCHOLchanges}) and order parameters
in equimolar mixtures (Fig. \ref{OrderParametersCHOL}) are overestimated by all force fields.
%For Berger/H\"oltje and MACROG models the overestimation is significant,
%while less severe for CHARMM36 and Slipid models.
The overestimation of order parameters observed in CHARMM without cholesterol and in all force fields
with cholesterol is smaller than the contribution by undulations in large simulations (section \ref{undulations} in the supplementary information).
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/slopes.eps}
  \caption{\label{OrderParametersCHOLchanges}
    Slopes of order parameters as a function of cholesterol.
  }
\end{figure}

@@

\subsection{Lipid bilayer dimensions and density profiles as a function of cholesterol}

Lipid bilayer dimensions can be experimentally accessed by measuring X-ray scattering
form factor, which is related to the electron density along membrane normal
via Fourier transformation \cite{pan12,Heftberger15,Marquardt15,ollila16,??}.
%methods and have been widely used to validate lipid bilayer simulations \cite{ollila16,??}. 
The form factor can be translated to density profiles, 
%While sophisticated models are applied to extract
area per lipids and bilayer thickness using SDP model or its combination with MD simulations  \cite{pan12,Heftberger15,Marquardt15,??}.
Sophisticated approaches for single component lipid bilayers are available \cite{??},
but multicomponent systems are more difficult to interpret \cite{pan12,Heftberger15,Marquardt15,??}.
%In such case the validation can be done by comparing directly measurable form factors between
%simulations and experiments. If simulation reproduces the form factor, it can be also
%used to interpret the experiments.

State of the art force fields give good agreement with experimental form factors for
pure POPC lipid bilayers \cite{??}, which is the case also here, except for the Berger
simulation (Fig.~\ref{FormFactors}).
%The form factors from different simulation models with
%different cholesterol concentrations are compared to experimental data in Fig.~\ref{FormFactors}. \\
Upon addition of cholesterol, the third minima in form factor systemically decreases from $\sim$0.42 to $\sim$0.32
%Cholesterol induced changes in lipid packing, thickness and density profiles are suggested
%to be relevant, for example, for membrane protein interactions and cellular membrane
%lateral organization \cite{??}. Experimental and simulation studies show that
%the cholesterol induces the so called ''condensing effect'', i.e. acyl chain ordering
%increases the membrane thickness and
%reduces the area per molecule in lipid bilayers \cite{??}.
This may be related to the reduced area per molecule upon addition of cholesterol (Fig. \ref{apls}).
%, showing the area per PC headgroups and per
%total number of molecules as a function of cholesterol concentration from different
%models.
The area per total amount of molecules decreas with cholesterol in all simulations, partly
due to the smaller area covered by the cholesterol than lipids, but also because
lipids ordered by cholesterol require less space. Due to the latter effect,
the area per PC headgroup does not essentially increase up to the addition of
$\sim$15 mol\% of cholesterol. 
\begin{figure}[]
 \centering
 \includegraphics[width=8cm]{../FIGS/FormFactors.eps}
 \caption{\label{FormFactors}
   Form factors from simulations and experiments.
 }
 \todo{Details about form factor calculation code is discussed in issues
   https://github.com/NMRLipids/MATCH/issues/56 and https://github.com/NMRLipids/MATCH/issues/50.
   Once the code is finalized, we should recalculate and check the form factors.
   CURRENTLY, 10\%-40\% cholesterol concetrations are calculated with a code which gives possibly incorrect
   heigths for the maxima.
 } \\
 \todo{The y-axis scale cannot be explicitly measured in experiments.
   Therefore, the y-axis of the experimental form factor is typically scaled to match with simulations.
   This currently not done, because we have several simulations which give inequal form factors, and therefore
   it is not clear against which simulation results we should scale the experimental results.
   I have created a issue for this discussion: https://github.com/NMRLipids/NmrLipidsCholXray/issues/18
 } \\
 \todo{Not all experimental and simulation data is here (maybe?).}
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/apls.eps}
  \caption{\label{apls}
    Area per molecules calculated from different simulation models
    as a function of cholesterol concentration. The solid symbols are area per
    total amount of molecules (chol+PC) and the empty symbols area per PC headgroups.
    Top figure shows absolute values and bottom figure shows changes respect
    to pure lipid system.
  }
\end{figure}

Visual inspection of form factors (Fig. \ref{FormFactors}) and quantitative quality estimation (Fig. \ref{FormFactorsFITNESS})
suggests that the Berge/Holtje force field gives the best agreement with experiments with large
cholesterol concentration. This is surprising because the same model overestimated acyl chain
order parameter increase upon addition of cholesterol more than CHARMM36 or Slipids simulations
(Figs. \ref{OrderParametersCHOL} and \ref{OrderParametersCHOLchanges}).
The Berger/Holtje model also exhibits the most pronounced decrease in the area per molecule
upon addition of cholesterol (Fig. \ref{alps}).
On the other hand, the electron density profiles from Berger/Holtje simulations are closer
to the result from the SDP model then other MD simulation models for both pure POPC bilayer
and equimolar POPC/cholesterol mixture, especially in the middle of the bilayer
where SDP model gives substantially lower density than MD simulations (Fig. \ref{densities}).
Therefore, the good quality of form factor Berger/Holtje model with large amounts of cholesterol
may rather indicate the importance of this low electron density in the bilayer center,
rather than better quality of lipid-cholesterol interactions in this model.
\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/FFfitness.eps}
  \caption{\label{FormFactorsFITNESS}
    Fitness of form factors between simulations and experiments.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=8cm]{../FIGS/densitiesEXTREMES.eps}
  \caption{\label{densities}
    Electron density profiles from MD simulations and SDP model.
  }
\end{figure}

@@

All the tested MD simulation models qualitatively reproduced experimentally observed condenced and
ordering effects of cholesterol. However, the acyl chain ordering was overestimated and quality
of form factors reduced with increasing amount of cholesterol. Although the biological importance
of these details is unclear, overestimated ordering may promote formation of liquid ordered phases and
inaccuate form factors complicate interpretation of scattering experiment with MD simulations.
Therefore, MD simulation force fields correctly reproducing the order effects upon addition of the
cholesterol would be invaluable.

To facilitate the comparison of the quality of complex lipid bilayers,
we introduce the quantitative quality measures which measure the quality of
intermolecular interactions in binary lipid bilayers from different perspectives.
The simple difference between experimental and simulated C-H bond order parameters
give a very detailed picture about the quality of molecular conformations in all
positions of lipid bilayer (Fig. \ref{OrderParametersCHOLfitness}).
However, complexity of such presentation increases with increasing amount of data
and it may be hard to present in practical format readable for human or machine.
Using root mean square deviation across regions in the molecule simplifies the presentation,
but compromizes the spatial resolution. Here, we divide the lipid into headgroup, glycerol
backbone, sn-1 and sn-2 chains to present quality of binary lipid bilayer in more convenient
format (Fig. \ref{OrderParametersCHOLfitness}).
\todo{Discussion to be finished once we have the fitness function defined.}

@@

%Cholesterol ordering effect is overestimated in Berger/Holtje and MacRog models.
%Slight overestimation is observed also in CHARMM36 and Slipid models, but more
%careful analysis is required to conclude if this is significant or not.
